Simple repression draft

### 2020-05-12 doc/paper.tex
Added much detail to means section, but more to do

%%%%% Referencing macros %%%%%
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\eqrange}[2]{Eq.~(\ref{#1}-\ref{#2})}

@@

\section{Introduction}
\mmnote{Key ideas, no particular order, that I haven't written down before:
\item Our goal is to build phenomenological models of input-output
functions of genetic circuits. We want concepts, not mere facts.
In particular a key question is: at this level of
coarse graining, what microscopic details do I need to explicitly
model, and how do we figure that out? For example, do I need to
worry about all or even any of the steps that individual RNAPs go
through each time they make a transcript? Turning the question
around, can we see any imprint of those processes in the
available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other
beautifully here.
\end{itemize}}
\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As shown in Figure~1.

@@

\item Change order in Fig~1 and order of presentation of noneq
models to match my treatment of Fano factors. This doesn't do
them in order of increasing math complexity, but it does do them
in logical order: Poisson, RNAP bound/unbound states, RNAP
unbound/closed/open states, ambiguous active/inactive, bursty
Poisson. For each model, sketch the intuition for its
construction and interpretation of the states/weights/rates,
possibly calling ahead/deferring some arguments to the
constitutive Fano section. Then quote the expression for mean
mRNA (that will be derived more carefully in an appendix), then
show the algebra to compute fold-change from mean mRNA and show
how it fits into form of the master curve.

@@

@@

@@

\includegraphics[width=\textwidth]{../figures/fig1/fig1.pdf}

@@

    (D) shows four possible coarse-graining in a nonequilibrium picture.

@@

>>>>

%%%%% Referencing macros %%%%%
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\tref}[1]{Table~\ref{#1}}
\newcommand{\eq}[1]{Eq.~(\ref{#1})}
\newcommand{\eqrange}[2]{Eqs.~(\ref{#1}-\ref{#2})}

@@

\section{Introduction}
\mmnote{Key ideas, no particular order, that I haven't written down before:
\begin{itemize}
\item Our goal is to build phenomenological models of input-output
functions of genetic circuits. More precisely, we want phenomenology
in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning, e.g., we don't want to coarse-grain
as far as Hill functions. Why not? We are motivated by studies
like~\cite{Razo-Mejia2020}, for which Hill functions are
insufficient. Such work requires a quantitative theory of how
biophysical changes at the molecular level propagate to
input-output functions at the genetic circuit level.
We want concepts, not mere facts.
In particular a key question is: at this level of
coarse graining, what microscopic details do I need to explicitly
model, and how do we figure that out? For example, do I need to
worry about all or even any of the steps that individual RNAPs go
through each time they make a transcript? Turning the question
around, can we see any imprint of those processes in the
available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling can complement each other
beautifully here.
\end{itemize}}

\fig{fig1:means_cartoons}A shows the qualitative picture of
simple repression. An operator, the binding site on the DNA for a
repressor protein, may be found occupied by a repressor, in which
case transcription is blocked from occurring. Or it may be found
unoccupied, in which case RNA polymerase (RNAP) may bind and
transcription can proceed.

The key assumption we make is that binding of repressor and RNAP in the promoter region of interest is exclusive, meaning that one or the other may bind, but never may both be simultaneously bound.
It is often imagined that when the repressor is bound to its operator,
RNAP is sterically blocked from binding to its promoter sequence.
Current evidence suggests this is sometimes, but not
always the case~\mmnote{Cite that fig, from ??Nathan or Bill or
Rob??, showing locations of repressor binding sites far upstream
of promoter}, and it remains an interesting open question precisely
how a repressor bound far upstream is able to repress transription.
Suggestions include ``action-at-a-distance'' mediated by kinks in
the DNA, formed when the repressor is bound, that prevent RNAP binding.
Nevertheless, our modeling in this work is sufficiently
coarse-grained that we simply assume exclusive binding and need
not worry about such precise details.

\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

\section{Means}
\subsection{Fold-changes are indistinguishable across models}
As a first stop on our search for the ``right'' model of simple
repression, let us consider what we can learn from theory and
measurements of fold-change in mean expression. The fold-change is defined as
\begin{equation}
\text{fold-change}
= \frac{\langle \text{gene expression with repressor present} \rangle}
        {\langle \text{gene expression with repressor absent} \rangle}
= \frac{\langle m\rangle(R\ne 0)}{\langle m\rangle(R = 0)}
\label{eq:fc_def}
\end{equation}
where angle brackets denote the averge over a population of cells
and mean mRNA $\langle m\rangle$ is viewed as a function of
repressor copy number $R$.
Also note the implicit assumption in equating ratios of gene
expression, i.e., protein levels, to mRNA levels. The second
equality in~\ref{eq:fc_def} amounts to an assumption that the
translation efficiency, i.e., the number of proteins translated
per mRNA, is the same in both conditions.
This is reasonable since the cells in the two conditions are
identical except for the presence of the transcription factor,
and it is difficult to imagine the transcription factor having
any direct effect on translation.

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with an without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.

Panels C and D of \fig{fig1:means_cartoons} depict a smorgasbord
of mathematized cartoons for simple repression that have appeared
in previous literature. For each cartoon, we calculate the
fold-change in mean gene expression as predicted by that model,
deferring some algebraic details to the appendix.
What we will find is that all cartoons collapse to a single
master curve, shown in \fig{fig1:means_cartoons}B, which contains
just two parameters. We label the parameters $\Delta F_R$, an
effective free energy parametrizing the repressor-DNA
interaction, and $\rho$, which subsumes all details of
transcription in the absence of repressors.
We will offer some intuition for why this master curve exists and discuss
why it is likely impossible to discriminate ``right'' from ``wrong''
cartoons given only measurements of fold-changes in expression.

\mmnote{Snappy names for all these cartoons would really help the reader. Also naming them by number of promoter states is confusing to the reader because analogous models have different number of states when we're talking about constitutive version vs version with repressor. Thoughts??}

\subsubsection{Equilibrium, 2-state}
In this simplest model, the promoter is idealized as existing in
one of two states, either repressor bound or repressor unbound.
The rate of transcription, and therefore the level of gene
expression, is assumed to be proportional to the fraction of time
spent in the repressor unbound state.
Mathematically, this amounts to the statement that the mean mRNA level is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound
state, $\gamma$ is the mRNA degradation rate, $R$ is repressor
copy number, $N_{NS}$ is the number of nonspecific binding sites
in the genome (where repressors spend most of their time when not
bound to the operator, their specifc binding site),
$\beta=1/k_BT$, and $\Delta\epsilon_R$ is the binding energy of a
repressor to its operator site. The derivation of this is deferred
to the appendix.

The fold-change is easily found as the ratio of mean mRNA with
and without repressor. This gives
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve
in~\eq{fig1:means_cartoons} with $\rho=1$.

In fact it was noted in~\cite{Chure2019} that this two-state
model can be viewed as the coarse-graining of any equilibrium
promoter model in which no transcriptionally active states have
transcription factor bound. Put differently, there is no overlap
between TF bound states and transcriptionally active states. We
will see this explicitly in the 3-state equilibrium model below,
but perhaps surprising is that something analogous carries over
even to the nonequilbrium models we consider later.

\subsubsection{Equilibrium, 3-state}
Compared to the previous model, here we fine-grain the repressor
unbound state into two separate states: empty, and RNAP bound.
This picture was used in~\cite{Garcia2011a} as we use it here,
and in~\cite{Razo-Mejia2018} and~\cite{Chure2019} it was generalized
to incorporate small-molecule inducers that bind the repressor.
Interestingly, the effect of this is simply to renormalize $R$
from the total number of repressors to a smaller effective number,
roughly (though not exactly) the number which are unbound by inducers.

The mean mRNA, as derived in the appendix from a similar
enumeration of states and weights as the previous model, is
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
\frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        },
\end{equation}
where the new variables are $\Delta\epsilon_P$, the difference in
RNAP binding energy to its specific site (the promoter) relative
to an average nonspecific background site, and the RNAP copy
number, $P$. The fold-change again follows immediately as
\begin{align}
\text{fold-change}
&= \frac{\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {
        1 + \frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}
        + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}
        }
\frac{1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
        {\frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\\
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\epsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\epsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R - \log\rho))^{-1},
\end{align}
with $\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})$
and $\rho = 1 + \frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$
as shown in~\fig{fig1:means_cartoons}.

\subsubsection{Nonequilibrium model 1 - Poisson promoter}
For our first kinetic model, we consider the simplest possible
picture with only two states, repressor bound and unbound.
This is exactly the model used for the main results of~\cite{Jones2014}.
In this picture, repressor association and dissociation rates
from its operator site, $k_R^+$ and $k_R^-$, respectively, govern
transitions between the two states. When the system is in the
unbound state, transcriptions initiate at rate $r$, which
represents a coarse-graining of all the downstream processes into
a single effective rate. mRNA is again degraded at rate $\gamma$.

Let $p_{m,R}(t)$ denote the joint probability of finding the
system in the repressor bound state with $m$ mRNA molecules
present at time $t$, and similarly define $p_{m,U}(t)$ for the
repressor unbound state. 
This model is governed by a master equation giving the time
evolution of $p_{m,R}(t)$ and $p_{m,U}(t)$, which is
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& - k_R^- p_{m,R}(t) + k_R^+ p_{m,U}(t)
                + (m+1)\gamma p_{m+1,R}(t) - \gamma p_{m,R}(t)
\\
\deriv{t}p_{m,U}(t) =&\; k_R^- p_{m,R}(t) - k_R^+ p_{m,U}(t)
                        + rp_{m-1,U}(t) - rp_{m,U}(t)
                \\
                &+ (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t).
\end{split}
\end{align}
In each equation, the first two terms describe transitions
between promoter states due to repressors unbinding and binding,
respectively. The final two terms describe degradation of mRNA,
decreasing the copy number by one, and the terms with coefficient
$r$ describe transcription initiation increasing the mRNA copy
number by one.

We can greatly simplify the notation, which will be especially
useful for the more complicated models yet to come, by
rexpressing the master equation in vector form. The promoter
states are collected into a vector and the rate constants are
collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix.
Taking steady state by setting time derivatives to zero,
the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to the appendix.
Recall $k_R^+$ is proportional to the repressor copy number, so
in computing fold-change, absence of repressor corresponds to
$k_R^+\rightarrow0$.
Therefore fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons} with $\rho=1$.

\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
Analogous to the three-state equilibrium model, this model
fine-grains the repressor unbound state of nonequilbrium model 1,
resolving it into an empty promoter state and an RNAP-bound state.
Note in this picture, in contrast with model 4 below,
transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and
empty states: if an RNAP successfully escapes the promoter and
proceeds to elongation of a transcript, clearly it is no longer
bound at the promoter! Therefore another RNAP must bind before
another transcript can be initiated.

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above.
The main subtlety arises since transcription initiation
accompanies a promoter state change, but this matrix can be
understood by analogy to the off-diagonal element of $\matr{K}$.
The master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_{in}} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_{in}} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & r \\ 0 & 0 & 0
                \end{pmatrix},\
\matr{R_{out}} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having
$m$ mRNA present along with the promoter having repressor bound,
being empty, or having RNAP bound, respectively.
$\matr{R_{in}}$ describes probability flux into the state
$\vect{p}_m$ from a state with one fewer mRNA, namely
$\vect{p}_{m-1}$, and $\matr{R_{out}}$ describes probability flux
out of the state $\vect{p}_m$ for a state with one more mRNA,
namely $\vect{p}_{m+1}$.
$\matr{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
\end{equation}
Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def},
from which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve with
$\rho = 1 + k_P^+/(k_P^- + r)$ as claimed.

\subsubsection{Nonequilibrium model 3
- Multistep transcription initiation and escape}
One might reasonably complain that the first two
``nonequilbrium'' models we have considered are straw men. Their
steady states necessarily satisfy detailed balance which is
equivalent to thermodynamic equilibrium. Why is this the case? At
steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a
linear chain, there is only one way in or out of the repressor
bound and RNAP bound states, implying each edge must actually
have a net zero probability flux, which is the definition of
detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

Now we consider model 3 in~\eq{fig1:means_cartoons} which allows
the possibility of true nonequilibrium steady-state fluxes
through the promoter states.
We point out that this model was considered previously in~\cite{Mitarai2015}
where a comparison was made with model 1 as used in~\cite{Jones2014}.
The authors of~\cite{Mitarai2015} argued that the additional
complexity is essential to properly account for the noise in the
mRNA distribution. We will weigh in on both models later when we
consider observables beyond fold-change.

The master equation governing this model is identical to model 2 above,
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_{out}} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_{in}} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with different matrices.
The state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound,
empty, RNAP closed complex, and RNAP open complex.
Besides increasing dimension by one again, the only new feature in
$\matr{K}$ is the rate $k_O$, representing the (irreversible)
rate of open complex formation from the closed complex.
The initiation matrices are given by
\begin{equation}
\matr{R_{in}} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & r \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_{out}} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0 \\ 0 & 0 & 0 & 0\\ 0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to model 2.

The expression for mean mRNA is substantially more complicated
now, as worked out in the appendix where we find
\begin{align}
\langle m\rangle &= \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)}
\\
&= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{align}
The strategy is to isolate the terms involving the repressor, so that now the fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve once again, with
$\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

This example hints that an arbitrarily fine-grained model of
downstream transcription steps may still be collapsed to the form
of the master curve in~\fig{fig1:means_cartoons}B so long as the
repressor binding is exclusive with transcriptionally active
states. We offer this as a conjecture, and we suspect that a
careful argument using the King-Altman diagram
method~\cite{King1956, Hill1966} might furnish a ``proof.''
Our focus here is not on full generality but rather to survey an
assortment of plausible models that have been proposed in the
literature for simple repression.

\subsubsection{Nonequilibrium model 4 - ``Active'' and ``inactive'' states}
This model is at the core of the theory in~\cite{Razo-Mejia2020}.
At a glance the cartoon for this model may appear very similar to model 2,
and mathematically it is, but the interpretation is rather different.
In model 2, we interpreted the third state literally as an
RNAP-bound promoter and modeled initiation of a transcript as
triggering a promoter state change, making the hopefully
uncontroversial assumption that an RNAP can only make one
transcript at a time. In contrast, in the present model the
promoter state does \textit{not} change when a transcript is
initiated. So we no longer interpret these states as literally
RNAP bound and unbound but instead as coarse-grained ``active''
and ``inactive'' states, the details of which we leave
unspecified for now. We will comment more on this model below
when we discuss Fano factors of models.

Mathematically this model is very similar to models 1 and 2.
Like model 1, the matrix $R$ describing transcription initiation
is diagonal, namely
\begin{equation}
\matr{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 0 & 0 & 0 \\ 0 & 0 & r
        \end{pmatrix}.
\end{equation}
so the master equation takes verbatim the same form as it did for model
1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
$K$ is the same as~\eq{eq:3state_cme_matrices_pt1} from model 2,
although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to
reiterate that these are not simply RNAP binding and unbinding rates.

Carrying out the algebra, the mean mRNA can be found to be
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
\frac{k_R^- k^+}
{k_R^- k^+ + k_R^- k^- + k_R^+ k^-},
\end{equation}
and the fold-change readily follows,
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k^+}{k_R^- k^+ + k_R^- k^- + k_R^+ k^-}
        \frac{k_R^- k^+ + k_R^- k^-}{k_R^- k^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \left(1 + \frac{k^+}{k^-}\right)^{-1}
        \right)^{-1},
\end{align}
from which we see $\rho = 1 + k^+/k^-$ as shown in~\fig{fig1:means_cartoons}.

\subsubsection{Nonequilibrium model 5 - Bursty promoter}
We will have much more to say about this model in subsequent sections.
For now we sketch it as an intuitive analog to model 1,
with just two states, repressor bound or unbound, and transition rates
between them of $k_R^+$ and $k_R^-$.
In model 1, when in the unbound state, single mRNA transcripts
are produced as a Poisson process with some characteristic rate
$r$. This model by contrast produces, at some Poisson rate $k_i$,
\textit{bursts} of transcripts with a geometrically distributed
burst size and a mean burst size $b$.

From this intuitive picture and by analogy to model 1, then, it should be plausible that the mean mRNA level is
\begin{equation}
\langle m\rangle = \frac{k_i b}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
which will turn out to be correct from a careful calculation.
For now, we simply note that just like model 1, the fold-change becomes
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1}
\end{equation}
with $\rho=1$ also like model 1.

We will also see later how this model emerges as a natural limit of model 4.

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}

@@

\item Key assumption: expression is proportional to occupation
probability of whatever the ``active'' state is for a given
model, whatever we mean by active.
\item Also emphasize that parameters/rates are \textit{not}
apples-to-apples between cartoons, especially the initiation rate $r$.
Indeed, this is the entire point! Except for repressor
energies/rates, the other parameters are not independently
measurable or inferrable b/c we're not even sure what
steps/states we should be writing in the model!
(simple RNAP kinetics as inferred in vitro, or supercoiling, or
RNAP cluster kinetics a la Ibrahim, or maybe
gene-copy/RNAP-copy/other-copy number is really the key, or...)
So all you get are coarse-grained combos like $\rho$, and even
$\rho$ would be hard to infer, much less trying to infer its innards.

@@

In theory you could refit their raw data to different models, but
as we see later, even having data of full mRNA distributions is
insufficient to identify parameters in all but the simplest of
the models in Figs.\ 1 and 2, so it seems doubtful that only
measurements of means would fare better.
??Double check, what did they actually measure in 2012 paper??

@@

\item Here's a fun puzzle: 2-state equilibrium model gets details
of transcription so wrong, how can it be useful? Because
fold-change is a ratio! Since details all collapse into $\rho$,
fitting this 2-state model to fold-change measurements can
\textit{still} give you a not-bad estimate of ratio of repressor
rates, i.e., binding energies.
\item 

@@

\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}

@@

    (D) shows five possible coarse-grainings in a nonequilibrium picture.

@@

\section{Discussion and future work}
\mmnote{
Do all my approximations work simply because galK happens to have
a low duty cycle of supercoiling, $\beta/\alpha$ in Chong2014
picture? A potentially interesting test: put in the dozen or so
Hernan constructs at some other locus, preferably one of the ones
Chong identified as having a large duty cycle (large $\beta/\alpha$),
i.e., the opposite limit of galK. Then just measure fold-changes.
You don't even need to do FISH, although that would be even
better. If their model is right, $\rho$ should be approximately
set by the duty cycle of supercoiling, NOT by RNAP kinetics.
\textit{This is a NICE prediction for future work.} \textit{cspE}
is claimed nonessential, prediction is you drop HG reporter
constructs in there and the inferred binding energies of all lacI
operators shift the same amount, by about $\ln(3)\sim 1kT$. Not
huge, but detectable, although might be confounding effects from
moving to a new locus. With FISH measurements, $k^+$ and $k^-$ in
Fig 2, model 4 should now be separately inferrable if Chong2014
measurements of $\beta/\alpha$ are correct. This would require an
obnoxious amount of cloning, but once you did that, the
measurements would be same old stuff that RP lab does.}
